---
title: "importMetadata"
author: "Melanie Smith"
date: "3 February 2025"
output: html_document
---


# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(tidyverse)
library(kableExtra)  # For pretty tables
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(pheatmap)

# set project directory
projectDir <- "/home/smit1924/foodFreq_miRNA_SCOPE_STOP"
dataFolder <- "/media/sf_D_DRIVE/VM_Projects/dataFolder"

expt_name <- 'foodFreq_miRNA_SCOPE_STOP'

SCOPE_dataDictionary_file <- file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/Data_dictionary_FINAL_DISTR_updated_130215_biomarker_SNP.xlsx")
SCOPE_raw_file <- file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/SCOPE_full_raw_20141201.rds")
STOP_raw_dat_file <- file.path(dataFolder,"STOPStudy_DATA_2021-04-23_1151.csv")
STOP_raw_dict_file <- file.path(dataFolder,"STOPStudy_DataDictionary_2021-04-23.csv")
SCOPE_STOP_BWcentile_metab_file <- file.path(dataFolder, "SCOPE_STOP_BWcentile_metab.csv")

SCOPE_STOP_full <- file.path(dataFolder, "SCOPE_STOP_full_20240208.rds")

outdir <- file.path(projectDir, paste0("output_", expt_name))

dir.create(outdir)

tertileTable_out_file <- file.path(outdir, "tertile_table.xlsx")
quartileTable_out_file <- file.path(outdir, "quartile_table.xlsx")

# REDCap import
source("/media/sf_D_DRIVE/VM_Projects/dataFolder/redcap_api_import.R")

# SCOPE samples with miRNA plasma
SCOPE_plasma <- c("SCP1220", "SCP1328","SCP1413", "SCP1414", "SCP1420", "SCP1421", "SCP1724", "SCP3412", "SCP3492", "SCP3580", "SCP3628", "SCP3637", "SCP3660", "SCP3738", "SCP3780", "SCP3825", "SCP3843", "SCP3847", "SCP3852", "SCP3872", "SCP3875", "SCP3877", "SCP3928", "SCP3929", "SCP3938", "SCP3940", "SCP3954", "SCP3962", "SCP3992", "SCP4010", "SCP4041", "SCP4059", "SCP4060", "SCP4073", "SCP4139", "SCP4148", "SCP4154", "SCP4157", "SCP4164", "SCP4177", "SCP4195", "SCP4196", "SCP4319", "SCP4378", "SCP4536", "SCP4538", "SCP4565", "SCP4578", "SCP4706", "SCP4726", "SCP4733", "SCP4748", "SCP4809", "SCP4913")

# STOP samples with miRNA plasma
STOP_plasma <- c("STP0016", "STP0023", "STP0105", "STP0148", "STP0230", "STP0275", "STP0282", "STP0378", "STP0582", "STP0583", "STP0593", "STP0596", "STP0636", "STP0677", "STP0692", "STP0761", "STP0788", "STP0795", "STP0798", "STP0852", "STP0867", "STP0888", "STP0903", "STP0912", "STP0932", "STP0944", "STP1002", "STP1082", "STP1087", "STP1206")

```

## Custom functions

```{r}
# This function performs recoding of categorical data (9-levels) in STOP to match (5-levels) in SCOPE.
# This allows for the SCOPE and STOP data to be combined
f10c.gps <- function(x){
  # Convert input x to a factor with levels 1 through 9
  # This ensures all values in x are mapped to these exact levels
  
  # First convert to numeric representation of the factor
  x_numeric <- as.numeric(x)
  
  x1 <- factor(x_numeric, 1:9)
  
  # Remap the original 9 levels to a new set of values:
  # - Original level 1 becomes 2
  # - Original level 2 becomes 3
  # - Original level 3 becomes 4
  # - Original level 4 becomes 4 (unchanged)
  # - Original level 5 becomes 5
  # - Original level 6 becomes 5
  # - Original level 7 becomes 5
  # - Original level 8 becomes 1
  # - Original level 9 becomes NA (missing value)
  levels(x1) <- c(2, 3, 4, 4, 5, 5, 5, 1, NA)
  
  # Convert to a factor again, but now with levels 1 through 5 only
  # This ensures the result only contains values 1-5
  # Any NA values remain as NA
  x1 <- factor(x1, 1:5)
  
  # Return the transformed factor
  return(x1)
}

# This function transforms categorical data into specific numeric values on a scale from 0 to 1
# This encodes the original categories (represented as frequencies) into a meaningful numeric scale
fastfd.fix <- function(x){
  # Convert input to a factor
  # This allows for categorical mapping of values
  x <- as.factor(x)
  
  # Replace the factor levels with specific numeric values
  # The values correspond to:
  # - First level: 0
  # - Second level: 0.1 (3/30)
  # - Third level: 0.214 (1.5/7)
  # - Fourth level: 0.643 (4.5/7)
  # - Fifth level: 1
  # These represent a non-linear scale from 0 to 1
  levels(x) <- c(0, 3/30, 1.5/7, 4.5/7, 1)
  
  # Convert factor to numeric values based on the new levels
  # as.matrix is used to ensure proper conversion
  x <- as.numeric(as.matrix(x))
  
  # Return the transformed numeric vector
  return(x)
}

# Function to count observations in each quartile for a variable
count_by_quartile <- function(variable_name, data) {
  # Skip if not numeric
  if(!is.numeric(data[[variable_name]])) return(NULL)
  
  # Get the variable data
  var_data <- data[[variable_name]]
  var_data <- na.omit(var_data)  # Remove NAs
  
  # Calculate quartiles
  quartiles <- quantile(var_data, probs = c(0, 0.25, 0.5, 0.75, 1), na.rm = TRUE)
  
  # Count observations in each quartile
  q1_count <- sum(var_data <= quartiles[2], na.rm = TRUE)
  q2_count <- sum(var_data > quartiles[2] & var_data <= quartiles[3], na.rm = TRUE)
  q3_count <- sum(var_data > quartiles[3] & var_data <= quartiles[4], na.rm = TRUE)
  q4_count <- sum(var_data > quartiles[4], na.rm = TRUE)
  
  # Return results
  data.frame(
    Variable = variable_name,
    Q1_threshold = quartiles[2],
    Q2_threshold = quartiles[3],
    Q3_threshold = quartiles[4],
    Q4_threshold = quartiles[5],
    Q1_count = q1_count,
    Q2_count = q2_count,
    Q3_count = q3_count,
    Q4_count = q4_count,
    Total_observations = length(var_data)
  )
}

# Function to count observations in each tertile for a variable
count_by_tertile <- function(variable_name, data) {
  # Skip if not numeric
  if (!is.numeric(data[[variable_name]])) return(NULL)
  
  # Get the variable data
  var_data <- data[[variable_name]]
  var_data <- na.omit(var_data)  # Remove NAs
  
  # Calculate tertiles
  tertiles <- quantile(var_data, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)
  
  # Count observations in each tertile
  t1_count <- sum(var_data <= tertiles[2], na.rm = TRUE)
  t2_count <- sum(var_data > tertiles[2] & var_data <= tertiles[3], na.rm = TRUE)
  t3_count <- sum(var_data > tertiles[3], na.rm = TRUE)
  
  # Return results
  data.frame(
    Variable = variable_name,
    T1_threshold = tertiles[2],  # 33rd percentile
    T2_threshold = tertiles[3],  # 66th percentile
    T3_threshold = tertiles[4],  # Max value
    T1_count = t1_count,         # Count in 1st tertile
    T2_count = t2_count,         # Count in 2nd tertile
    T3_count = t3_count,         # Count in 3rd tertile
    Total_observations = length(var_data)
  )
}


```


# Import the combined SCOPE and STOP data already processed
- These data come from work previously completed by Shalem.  
```{r}
# import Shalem's combined file
SCOPE_STOP_combined_20240208 <- readRDS(SCOPE_STOP_full) 

# create a new column with a samplename that matches the rest of my data
SCOPE_STOP_combined_20240208 <- SCOPE_STOP_combined_20240208 %>%
  mutate(samplename = case_when(
    Study == "SCOPE" ~ paste0("SCP", str_pad(Study.ID, width = 4, pad = "0")),
    Study == "STOP" ~ paste0("STP", str_pad(Study.ID, width = 4, pad = "0")),
    TRUE ~ NA_character_
  )) %>%
  dplyr::select(., samplename, everything())

# subset the combined file to only include the samples we have miRseq plasma for
subset_SCOPE_STOP <- subset(SCOPE_STOP_combined_20240208, samplename %in% c(SCOPE_plasma, STOP_plasma)) %>%
  # remove the columns we aren't interested in
  dplyr::select(., samplename, f1_age, BMI, SEI, f11_waist, GA_samp, Birthweight, Birthweight_centile, fetal_sex, GA_birth, Uncomplicated, FastFd_PrePregWK, FastFd_1stVstWK)

# Add a label to the 'GA_samp' column
attr(subset_SCOPE_STOP$GA_samp, "label") <- "Gestational age at sample collection (weeks)"

```

# Import SCOPE
```{r}

# import the full SCOPE metadata file
SCOPE_raw <- readRDS(SCOPE_raw_file)

# import SCOPE full data dictionary
SCOPE_dataDictionary <- read_excel(SCOPE_dataDictionary_file,
                                   sheet = "Stage 1 Data") %>%
  as.data.frame()

# # date variables for bloods and birth
# date_variables <- c("f1_date_1st_scope_vst", "f11_serum_date", "f11_edta_plasma_date", "f20_edta_plasma_date", "f23_date_20w_scan", "f39c_final_del_gest", "f26_Baby_DOB")
# date_SCOPE_dataDictionary <- dplyr::filter(SCOPE_dataDictionary, Variable_SAS_name %in% date_variables)%>%
#   dplyr::select(., Variable_SAS_name, "Variable Explanation\r\n", `Database Categories`)
# 
# # make a vector of interesting variables (food frequency questionnaire)
# fastFood_variables <- c("f10_burger_1st_vst", "f10_burger_1m_prepreg", "f19_burger_2nd_vst","f19_curry_2nd_vst", "f10_curry_1st_vst", "f10_curry_1m_prepreg", "f19_fried_chicken_2nd_vst", "f10_fried_chicken_1st_vst", "f10_fried_chicken_1m_prepreg", "f19_fruit_2nd_vst", "f10_fruit_1st_vst", "f10_fruit_1m_prepreg", "f19_leafy_vege_2nd_vst", "f10_leafy_vege_1st_vst", "f10_leafy_vege_1m_prepreg", "f19_chips_2nd_vst", "f10_chips_1st_vst", "f10_chips_1m_prepreg", "f19_oily_fish_2nd_vst", "f10_oily_fish_1st_vst", "f10_oily_fish_1m_prepreg", "f19_oth_fish_2nd_vst", "f10_oth_fish_1st_vst", "f10_oth_fish_1m_prepreg", "f19_pizza_2nd_vst", "f10_pizza_1st_vst", "f10_pizza_1m_prepreg", "f10_multivit_1st_vst", "f19_multivit_2nd_vst", "f10_multivit_1st_trim", "f10_multivit_prepreg", "f10c_burger_1st_vst", "f10c_burger_1m_prepreg", "f19c_burger_2nd_vst", "f19c_curry_2nd_vst", "f10c_curry_1st_vst", "f10c_curry_1m_prepreg", "f19c_fried_chicken_2nd_vst", "f10c_fried_chicken_1st_vst", "f10c_fried_chicken_1m_prepreg", "f19c_fruit_2nd_vst", "f10c_fruit_1st_vst", "f10c_fruit_1st_vst_4gp", "f10c_fruit_1m_prepreg", "f19c_leafy_vege_2nd_vst", "f10c_leafy_vege_1st_vst", "f10c_leafy_vege_1m_prepreg", "f19c_chips_2nd_vst", "f10c_chips_1st_vst", "f10c_chips_1m_prepreg", "f19c_oily_fish_2nd_vst", "f10c_oily_fish_1st_vst", "f10c_oily_fish_1m_prepreg", "f19c_oth_fish_2nd_vst", "f10c_oth_fish_1st_vst", "f10c_oth_fish_1m_prepreg", "f19c_pizza_2nd_vst", "f10c_pizza_1st_vst", "f10c_pizza_1m_prepreg")
# 
# fastFood_SCOPE_dataDictionary <- dplyr::filter(SCOPE_dataDictionary, Variable_SAS_name %in% fastFood_variables)%>%
#   dplyr::select(., Variable_SAS_name, "Variable Explanation\r\n", `Database Categories`)
# 
# # maternal and birth variables
# maternal_variables <- c("f11_wgt", "f11_bmi", "f11_hgt", "f11_waist", "f11_hip", "f26_Sex", "f1_age",  "f26_Birthwgt")
# 
# maternal_SCOPE_dataDictionary <- dplyr::filter(SCOPE_dataDictionary, Variable_SAS_name %in% maternal_variables)%>%
#   dplyr::select(., Variable_SAS_name, "Variable Explanation\r\n", `Database Categories`)

# putting together only the variables we need for Anand (not including the one's from Shalem's table)
final_variables <- c("f10c_fruit_1st_vst", "f10c_fruit_1m_prepreg", "f10c_leafy_vege_1st_vst", "f10c_leafy_vege_1m_prepreg", "f10c_oily_fish_1st_vst", "f10c_oily_fish_1m_prepreg")


# subset the SCOPE data dictionary so I don't have to keep scrolling for the variable explanations
subsetSCOPE_dataDict <- dplyr::filter(SCOPE_dataDictionary, Variable_SAS_name %in% final_variables) %>%
  dplyr::select(., Variable_SAS_name, "Variable Explanation\r\n", `Database Categories`)

# subset to the required columns
subsetSCOPE <- dplyr::select(SCOPE_raw, regid, centre, all_of(final_variables)) %>% 
  # remove all centres except Adelaide
  dplyr::filter(., centre == "Adelaide University") %>% 
    # sort by ID
  dplyr::arrange(., regid) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0("SCP", stringr::str_pad(regid, width=4, pad="0"))) %>%
  # make samplename the first column and drop column 'centre'
  dplyr::select(., samplename, everything(), -centre) %>% 
  # replace all the missing data codes with NA
  mutate(across(everything(), function(x){replace(x, which(x<0), NA)})) %>%
  dplyr::select(., -regid)
head(subsetSCOPE)
dim(subsetSCOPE)

# filter the metadata to only inlcude SCOPE samples with plasma miRNA
subsetSCOPE %<>% dplyr::filter(., samplename %in% SCOPE_plasma)

# # ensure R sees all the date columns as dates
# # Load necessary libraries
# library(dplyr)
# library(lubridate)
# 
# # Get names of date columns (all columns containing "date" or "DOB" in their names)
# date_cols <- names(subsetSCOPE)[grepl("date|DOB", names(subsetSCOPE), ignore.case = TRUE)]
# 
# # Convert each date column
# subsetSCOPE_date <- subsetSCOPE %>%
#   mutate(across(all_of(date_cols), 
#                 ~as.Date(as.character(.), 
#                          format = "%d/%m/%Y")))
# 
# # Verify the conversion worked correctly
# str(subsetSCOPE_date[, date_cols])
```

## Convert SCOPE variables to match STOP
- Fetal sex is coded as 1/2 in SCOPE; convert to 'Male' and 'Female' to match STOP.  

```{r}
# # Convert the 1/2 sex code to Male/Female
# subsetSCOPE %<>% dplyr::mutate(f26_Sex_MF = case_when(
#     f26_Sex == 1 ~ "Male",
#     f26_Sex == 2 ~ "Female"
#   ))

```

- Now convert to a frequency metric 

```{r}
# use the `fastfd.fix()` function to convert the catagorical variable to a fractional variable representing serves per day
# prepreg
subsetSCOPE$oilyFishPrePreg <- fastfd.fix(subsetSCOPE$f10c_oily_fish_1m_prepreg)
subsetSCOPE$fruitPrePreg <- fastfd.fix(subsetSCOPE$f10c_fruit_1m_prepreg)
subsetSCOPE$leafyVegPrePreg <- fastfd.fix(subsetSCOPE$f10c_leafy_vege_1m_prepreg)

# 1st-trim
subsetSCOPE$oilyFish1stVisit <- fastfd.fix(subsetSCOPE$f10c_oily_fish_1st_vst)
subsetSCOPE$fruit1stVisit <- fastfd.fix(subsetSCOPE$f10c_fruit_1st_vst)
subsetSCOPE$leafyVeg1stVisit <- fastfd.fix(subsetSCOPE$f10c_leafy_vege_1st_vst)

# multiply by 7 to calculate the serves per week
# prepreg
subsetSCOPE$oilyFishPrePregWK <- subsetSCOPE$oilyFishPrePreg*7
subsetSCOPE$fruitPrePregWK <- subsetSCOPE$fruitPrePreg*7
subsetSCOPE$leafyVegPrePregWK <- subsetSCOPE$leafyVegPrePreg*7

# 1st-trim
subsetSCOPE$oilyFish1stVisitWK <- subsetSCOPE$oilyFish1stVisit*7
subsetSCOPE$fruit1stVisitWK <- subsetSCOPE$fruit1stVisit*7
subsetSCOPE$leafyVeg1stVisitWK <- subsetSCOPE$leafyVeg1stVisit*7
```



# Import STOP

```{r import STOP}

# STOP (REDCap import)
STOP_raw_dat <- read.csv("/media/sf_D_DRIVE/VM_Projects/dataFolder/STOPStudy_DATA_2021-04-23_1151.csv",
                         header = TRUE)
STOP_raw_dict <- read.csv(file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/STOPStudy_DataDictionary_2021-04-23.csv"
  ),
                          header = TRUE)

STOP_raw <- apply_data_dict(STOP_raw_dat,STOP_raw_dict)
table(STOP_raw$final_data) #use final data subset (N=1300) only

# variables I want
STOP_variables <- c("participant_id", "f8_fruit_1st_trim", "f8_fruit_prepreg", "f8_leafy_vege_1st_trim", "f8_leafy_vege_prepreg", "f8_oily_fish_1st_trim", "f8_oily_fish_prepreg")

# subset for the variables we want
subsetSTOP <- dplyr::select(STOP_raw, final_data, all_of(STOP_variables)) %>%
  # drop samples not in the final data subset
  dplyr::filter(., final_data == "Yes") %>% 
  # sort by ID
  dplyr::arrange(., participant_id) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(participant_id, width = 4, pad = "0"))
    ) %>%
  # make samplename the first column
  dplyr::select(., samplename, everything()) %>%
  dplyr::select(., -final_data, -participant_id)

# filter the metadata to only inlcude STOP samples with plasma miRNA
subsetSTOP %<>% dplyr::filter(., samplename %in% STOP_plasma)

# quick double check to see the function works as expected
temp <- data.frame(old = subsetSTOP$f8_leafy_vege_1st_trim,
                   numeric = as.numeric(subsetSTOP$f8_leafy_vege_1st_trim),
                   new = f10c.gps(subsetSTOP$f8_leafy_vege_1st_trim))

```

## Convert STOP variables to match SCOPE
- Height is measured in m. Recode to cm to match all the other measurements and SCOPE.    

```{r}
# # Convert the height in M to height in cm
# subsetSTOP %<>% dplyr::mutate(f9_hgt_cm = f9_hgt * 100)

```

- STOP variables have 9 catagorical groups, SCOPE has 5  
- Using the `f10c.gps` funciton, recode STOP to match SCOPE f10c_ variables  

```{r}
## regroup vars
# prepreg - I'm using the same naming convention here that Shalem used in his code
# basically this first step (prior to fastfd.fix() converts the 9 STOP cats to match the 5 SCOPE cats so I use the SCOPE "f10C_" name
subsetSTOP$f10c_oily_fish_1m_prepreg <- f10c.gps(subsetSTOP$f8_oily_fish_prepreg)
subsetSTOP$f10c_fruit_1m_prepreg <- f10c.gps(subsetSTOP$f8_fruit_prepreg)
subsetSTOP$f10c_leafy_vege_1m_prepreg <- f10c.gps(subsetSTOP$f8_leafy_vege_prepreg)

# 1st-trim
subsetSTOP$f10c_oily_fish_1st_vst <- f10c.gps(subsetSTOP$f8_oily_fish_1st_trim)
subsetSTOP$f10c_fruit_1st_vst <- f10c.gps(subsetSTOP$f8_fruit_1st_trim)
subsetSTOP$f10c_leafy_vege_1st_vst <- f10c.gps(subsetSTOP$f8_leafy_vege_1st_trim)
```

- Now convert to a frequency metric  

```{r}
# use the `fastfd.fix()` function to convert the catagorical variable to a fractional variable representing serves per day
# prepreg
subsetSTOP$oilyFishPrePreg <- fastfd.fix(subsetSTOP$f10c_oily_fish_1m_prepreg)
subsetSTOP$fruitPrePreg <- fastfd.fix(subsetSTOP$f10c_fruit_1m_prepreg)
subsetSTOP$leafyVegPrePreg <- fastfd.fix(subsetSTOP$f10c_leafy_vege_1m_prepreg)

# 1st-trim
subsetSTOP$oilyFish1stVisit <- fastfd.fix(subsetSTOP$f10c_oily_fish_1st_vst)
subsetSTOP$fruit1stVisit <- fastfd.fix(subsetSTOP$f10c_fruit_1st_vst)
subsetSTOP$leafyVeg1stVisit <- fastfd.fix(subsetSTOP$f10c_leafy_vege_1st_vst)

# multiply by 7 to calculate the serves per week
# prepreg
subsetSTOP$oilyFishPrePregWK <- subsetSTOP$oilyFishPrePreg*7
subsetSTOP$fruitPrePregWK <- subsetSTOP$fruitPrePreg*7
subsetSTOP$leafyVegPrePregWK <- subsetSTOP$leafyVegPrePreg*7

# 1st-trim
subsetSTOP$oilyFish1stVisitWK <- subsetSTOP$oilyFish1stVisit*7
subsetSTOP$fruit1stVisitWK <- subsetSTOP$fruit1stVisit*7
subsetSTOP$leafyVeg1stVisitWK <- subsetSTOP$leafyVeg1stVisit*7

```

# Combine SCOPE and STOP

```{r combine scope and stop}

combinedfruitFishLeafyVeg <- data.frame(samplename = c(subsetSCOPE[,"samplename"], subsetSTOP[,"samplename"]),
                               cohort = c(rep("SCOPE", nrow(subsetSCOPE)), rep("STOP", nrow(subsetSTOP))),
                               
                               oilyFishPrePregWK = c(subsetSCOPE[,"oilyFishPrePregWK"], subsetSTOP[,"oilyFishPrePregWK"]),
                               oilyFish1stVisitWK = c(subsetSCOPE[,"oilyFish1stVisitWK"], subsetSTOP[,"oilyFish1stVisitWK"]),
                               fruitPrePregWK = c(subsetSCOPE[,"fruitPrePregWK"], subsetSTOP[,"fruitPrePregWK"]),
                               fruit1stVisitWK = c(subsetSCOPE[,"fruit1stVisitWK"], subsetSTOP[,"fruit1stVisitWK"]),
                               leafyVegPrePregWK = c(subsetSCOPE[,"leafyVegPrePregWK"], subsetSTOP[,"leafyVegPrePregWK"]),
                               leafyVeg1stVisitWK = c(subsetSCOPE[,"leafyVeg1stVisitWK"], subsetSTOP[,"leafyVeg1stVisitWK"])
                               )
```

# Combine the data frames

```{r}
# put the SCOPE and STOP information together
finalTable <- dplyr::left_join(subset_SCOPE_STOP, combinedfruitFishLeafyVeg, by = "samplename")
# Some of the columns have label attributes and I want to get rid of them for the summary tables to work
# check which columns have label attributes
sapply(finalTable, attr, "label")
# remove labels where they exist
finalTable_stripped <- finalTable %>%
  mutate(across(everything(), ~{ attr(.x, "label") <- NULL; .x }))
# check label attributes have been removed
sapply(finalTable_stripped, attr, "label")

# Convert all labelled variables to numeric
for(col in names(finalTable_stripped)) {
  if("labelled" %in% class(finalTable_stripped[[col]])) {
    finalTable_stripped[[col]] <- as.numeric(finalTable_stripped[[col]])
  }
}

# summarise male/female + complicated/uncomplicated
tempTable <- finalTable_stripped %>%
  mutate(uncomplicated = case_when(
    Uncomplicated == 1 ~ "No",
    Uncomplicated == 2 ~ "Yes",
    TRUE ~ as.character(Uncomplicated)
  ),
  sex = case_when(
    fetal_sex == 1 ~ "male",
    fetal_sex == 2 ~ "female",
    TRUE ~ as.character(fetal_sex)
  )) %>%
  dplyr::select(., samplename, uncomplicated, sex) %>%
  dplyr::left_join(., SCOPE_STOP_combined_20240208[, c("samplename", "Outcome", "GDMn", "PEn", "PTBn", "SGAn", "Ethnicity")], by = "samplename")
# cross tabulation of complication status and fetal sex
table(tempTable$sex, tempTable$Outcome)

library(purrr)

# Mutate to create "finalOutcome"
tempTable <- tempTable %>%
  mutate(finalOutcome = pmap_chr(select(., GDMn, PEn, PTBn, SGAn), function(...) {
    values <- c(...)   # Get column values
    cols_with_yes <- names(values)[values == "Yes"]  # Get column names where value is "Yes"
    
    if (length(cols_with_yes) == 0) {
      return("uncomplicated")  # If no "Yes", mark as "uncomplicated"
    } else {
      return(paste(cols_with_yes, collapse = ", "))  # Concatenate column names with ", "
    }
  }))

# Print the result
print(df)
# cross tabulation of complication status and fetal sex
table(tempTable$sex, tempTable$finalOutcome)

```

## Generate quartile summary table

```{r}

# Quartile Summary Table
quartile_summary <- finalTable_stripped %>%
  summarise(
    across(c(f1_age, BMI, SEI, GA_samp, Birthweight, Birthweight_centile, FastFd_PrePregWK, FastFd_1stVstWK, 
             oilyFishPrePregWK, oilyFish1stVisitWK, fruitPrePregWK, fruit1stVisitWK, leafyVegPrePregWK, leafyVeg1stVisitWK),
           list(
             Min = ~min(.x, na.rm = TRUE),
             Q1 = ~quantile(.x, 0.25, na.rm = TRUE),
             Median = ~median(.x, na.rm = TRUE),
             Q3 = ~quantile(.x, 0.75, na.rm = TRUE),
             Max = ~max(.x, na.rm = TRUE),
             Mean = ~mean(.x, na.rm = TRUE),
             SD = ~sd(.x, na.rm = TRUE)
           )
    )
  ) %>%
  tidyr::pivot_longer(cols = everything(), 
               names_to = c("Variable", "Statistic"), 
               names_pattern = "(.+)_(Min|Q1|Median|Q3|Max|Mean|SD)", # Fix splitting
               values_to = "Value") %>%
  arrange(Variable, Statistic)

print(quartile_summary)


# Reshape the data to have variables as rows and statistics as columns
publication_table <- quartile_summary %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  ) %>%
  # Rearrange columns in a logical order
  select(Variable, Min, Q1, Median, Mean, Q3, Max)

# Create a publication-ready table
publication_table %>%
  kable(format = "html", 
        caption = "Summary Statistics by Variable",
        digits = 1) %>%  # Round to 1 decimal place
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)  # Bold header row

# Apply function to all numeric variables
quartile_distribution <- lapply(names(finalTable_stripped), function(var) {
  count_by_quartile(var, finalTable_stripped)
}) %>% bind_rows()

# View the results
print(quartile_distribution)

# Join with your original publication table
publication_table_quartile_complete <- publication_table %>%
  left_join(quartile_distribution %>% 
              select(Variable, Q1_count, Q2_count, Q3_count, Q4_count, Total_observations),
            by = "Variable")

# Display the enhanced table
publication_table_quartile_complete %>%
  kable(format = "html", 
        caption = "Summary Statistics and Quartile Distribution by Variable",
        digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# To save the table

write_xlsx(publication_table_quartile_complete, quartileTable_out_file)

```

## Generate Tertile summary table

```{r}

# Tertile Summary Table
tertile_summary <- finalTable_stripped %>%
  summarise(
    across(c(f1_age, BMI, SEI, GA_samp, Birthweight, Birthweight_centile, FastFd_PrePregWK, FastFd_1stVstWK, 
             oilyFishPrePregWK, oilyFish1stVisitWK, fruitPrePregWK, fruit1stVisitWK, leafyVegPrePregWK, leafyVeg1stVisitWK),
           list(
             Min = ~min(.x, na.rm = TRUE),
             T1 = ~quantile(.x, 1/3, na.rm = TRUE),  # 33rd percentile
             Median = ~median(.x, na.rm = TRUE),     # 50th percentile
             T2 = ~quantile(.x, 2/3, na.rm = TRUE),  # 66th percentile
             Max = ~max(.x, na.rm = TRUE),
             Mean = ~mean(.x, na.rm = TRUE),
             SD = ~sd(.x, na.rm = TRUE)
           )
    )
  ) %>%
  pivot_longer(cols = everything(), 
               names_to = c("Variable", "Statistic"), 
               names_pattern = "(.+)_(Min|T1|Median|T2|Max|Mean|SD)", # Update names_pattern
               values_to = "Value") %>%
  arrange(Variable, Statistic)

# Print the summary
print(tertile_summary)

# Reshape the data to have variables as rows and statistics as columns
publication_table_tertile <- tertile_summary %>%
  pivot_wider(
    names_from = Statistic,
    values_from = Value
  ) %>%
  # Rearrange columns in a logical order
  select(Variable, Min, T1, Median, Mean, T2, Max)

# Create a publication-ready table
publication_table_tertile %>%
  kable(format = "html", 
        caption = "Summary Statistics by Variable",
        digits = 1) %>%  # Round to 1 decimal place
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)  # Bold header row

# Apply function to all numeric variables
tertile_distribution <- lapply(names(finalTable_stripped), function(var) {
  count_by_tertile(var, finalTable_stripped)
}) %>% bind_rows()

# View the results
print(tertile_distribution)

# Join with your original publication table
publication_table_tertile_complete <- publication_table_tertile %>%
  left_join(tertile_distribution %>% 
              select(Variable, T1_count, T2_count, T3_count, Total_observations),
            by = "Variable")

# Display the enhanced table
publication_table_tertile_complete %>%
  kable(format = "html", 
        caption = "Summary Statistics and Tertile Distribution by Variable",
        digits = 1) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)

# To save the table
library(writexl)
write_xlsx(publication_table_tertile_complete, tertileTable_out_file)

```
