---
title: "importMetadata"
author: "Melanie Smith"
date: "3 February 2025"
output: html_document
---


# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(pheatmap)

# set project directory
projectDir <- "/home/smit1924/foodFreq_miRNA_SCOPE_STOP"
dataFolder <- "/home/smit1924/dataFolder"

expt_name <- 'foodFreq_miRNA_SCOPE_STOP'

SCOPE_dataDictionary_file <- file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/Data_dictionary_FINAL_DISTR_updated_130215_biomarker_SNP.xlsx")
SCOPE_raw_file <- file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/SCOPE_full_raw_20141201.rds")
STOP_raw_dat_file <- file.path(dataFolder,"STOPStudy_DATA_2021-04-23_1151.csv")
STOP_raw_dict_file <- file.path(dataFolder,"STOPStudy_DataDictionary_2021-04-23.csv")
SCOPE_STOP_BWcentile_metab_file <- file.path(dataFolder, "SCOPE_STOP_BWcentile_metab.csv")

outdir <- file.path(projectDir, paste0("output_", expt_name))

dir.create(outdir)

# REDCap import
source("/media/sf_D_DRIVE/VM_Projects/dataFolder/redcap_api_import.R")

```

# Import SCOPE
```{r}

# import the full SCOPE metadata file
SCOPE_raw <- readRDS(SCOPE_raw_file)

# import SCOPE full data dictionary
SCOPE_dataDictionary <- read_excel(SCOPE_dataDictionary_file,
                                   sheet = "Stage 1 Data") %>%
  as.data.frame()

# make a vector of interesting variables (food frequency questionnaire)
SCOPE_variables <- c("f11_wgt", "f11_bmi", "f11_hgt", "f11_waist", "f11_hip", "f26_Sex", "f1_age", "f39c_final_del_gest", "f26_Birthwgt",  "f10_burger_1st_vst", "f10_burger_1m_prepreg", "f19_burger_2nd_vst","f19_curry_2nd_vst", "f10_curry_1st_vst", "f10_curry_1m_prepreg", "f19_fried_chicken_2nd_vst", "f10_fried_chicken_1st_vst", "f10_fried_chicken_1m_prepreg", "f19_fruit_2nd_vst", "f10_fruit_1st_vst", "f10_fruit_1m_prepreg", "f19_leafy_vege_2nd_vst", "f10_leafy_vege_1st_vst", "f10_leafy_vege_1m_prepreg", "f19_chips_2nd_vst", "f10_chips_1st_vst", "f10_chips_1m_prepreg", "f19_oily_fish_2nd_vst", "f10_oily_fish_1st_vst", "f10_oily_fish_1m_prepreg", "f19_oth_fish_2nd_vst", "f10_oth_fish_1st_vst", "f10_oth_fish_1m_prepreg", "f19_pizza_2nd_vst", "f10_pizza_1st_vst", "f10_pizza_1m_prepreg", "f10_multivit_1st_vst", "f19_multivit_2nd_vst", "f10_multivit_1st_trim", "f10_multivit_prepreg", "f10c_burger_1st_vst", "f10c_burger_1m_prepreg", "f19c_burger_2nd_vst", "f19c_curry_2nd_vst", "f10c_curry_1st_vst", "f10c_curry_1m_prepreg", "f19c_fried_chicken_2nd_vst", "f10c_fried_chicken_1st_vst", "f10c_fried_chicken_1m_prepreg", "f19c_fruit_2nd_vst", "f10c_fruit_1st_vst", "f10c_fruit_1st_vst_4gp", "f10c_fruit_1m_prepreg", "f19c_leafy_vege_2nd_vst", "f10c_leafy_vege_1st_vst", "f10c_leafy_vege_1m_prepreg", "f19c_chips_2nd_vst", "f10c_chips_1st_vst", "f10c_chips_1m_prepreg", "f19c_oily_fish_2nd_vst", "f10c_oily_fish_1st_vst", "f10c_oily_fish_1m_prepreg", "f19c_oth_fish_2nd_vst", "f10c_oth_fish_1st_vst", "f10c_oth_fish_1m_prepreg", "f19c_pizza_2nd_vst", "f10c_pizza_1st_vst", "f10c_pizza_1m_prepreg")

# subset the SCOPE data dictionary so I don't have to keep scrolling for the variable explanations
subsetSCOPE_dataDict <- dplyr::filter(SCOPE_dataDictionary, Variable_SAS_name %in% SCOPE_variables) %>%
  dplyr::select(., Variable_SAS_name, "Variable Explanation\r\n", `Database Categories`)

# subset to the required columns
subsetSCOPE <- dplyr::select(SCOPE_raw, regid, centre, all_of(SCOPE_variables)) %>% 
  # remove all centres except Adelaide
  dplyr::filter(., centre == "Adelaide University") %>% 
    # sort by ID
  dplyr::arrange(., regid) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0("SCP", stringr::str_pad(regid, width=4, pad="0"))) %>%
  # make samplename the first column and drop column 'centre'
  dplyr::select(., samplename, everything(), -centre) %>% 
  # replace all the missing data codes with NA
  mutate(across(everything(), function(x){replace(x, which(x<0), NA)}))
head(subsetSCOPE)
dim(subsetSCOPE)

# SCOPE samples with miRNA plasma
SCOPE_plasma <- c("SCP1220", "SCP1328","SCP1413", "SCP1414", "SCP1420", "SCP1421", "SCP1724", "SCP3412", "SCP3492", "SCP3580", "SCP3628", "SCP3637", "SCP3660", "SCP3738", "SCP3780", "SCP3825", "SCP3843", "SCP3847", "SCP3852", "SCP3872", "SCP3875", "SCP3877", "SCP3928", "SCP3929", "SCP3938", "SCP3940", "SCP3954", "SCP3962", "SCP3992", "SCP4010", "SCP4041", "SCP4059", "SCP4060", "SCP4073", "SCP4139", "SCP4148", "SCP4154", "SCP4157", "SCP4164", "SCP4177", "SCP4195", "SCP4196", "SCP4319", "SCP4378", "SCP4536", "SCP4538", "SCP4565", "SCP4578", "SCP4706", "SCP4726", "SCP4733", "SCP4748", "SCP4809", "SCP4913")

# filter the metadata to only inlcude SCOPE samples with plasma miRNA
subsetSCOPE %<>% dplyr::filter(., samplename %in% SCOPE_plasma)


```

## Convert SCOPE variables to match STOP
- Fetal sex is coded as 1/2 in SCOPE; convert to 'Male' and 'Female' to match STOP.  

```{r}
# Convert the 1/2 sex code to Male/Female
subsetSCOPE %<>% dplyr::mutate(f26_Sex_MF = case_when(
    f26_Sex == 1 ~ "Male",
    f26_Sex == 2 ~ "Female"
  ))

```


# Import STOP

```{r import STOP}

# STOP (REDCap import)
STOP_raw_dat <- read.csv("/media/sf_D_DRIVE/VM_Projects/dataFolder/STOPStudy_DATA_2021-04-23_1151.csv",
                         header = TRUE)
STOP_raw_dict <- read.csv(file.path("/media/sf_D_DRIVE/VM_Projects/dataFolder/STOPStudy_DataDictionary_2021-04-23.csv"
  ),
                          header = TRUE)

STOP_raw <- apply_data_dict(STOP_raw_dat,STOP_raw_dict)
table(STOP_raw$final_data) #use final data subset (N=1300) only

# variables I want
STOP_variables <- c("participant_id", "f1_age", "f9_wgt", "f9_bmi", "f9_hgt", "f9_waist", "f9_hip", "f11_age", "f22_sex", "f22_gestw_del", "f22_birthwgt", "f8_burgers_1st_trim", "f8_burgers_prepreg", "f8_curry_1st_trim", "f8_curry_prepreg", "f8_chicken_1st_trim", "f8_chicken_prepreg", "f8_fruit_1st_trim", "f8_fruit_prepreg", "f8_leafy_vege_1st_trim", "f8_leafy_vege_prepreg", "f8_chips_1st_trim", "f8_chips_prepreg", "f8_oily_fish_1st_trim", "f8_oily_fish_prepreg", "f8_oth_fish_1st_trim", "f8_oth_fish_prepreg", "f8_pizza_1st_trim", "f8_pizza_prepreg")

# subset for the variables we want
subsetSTOP <- dplyr::select(STOP_raw, final_data, all_of(STOP_variables)) %>%
  # drop samples not in the final data subset
  dplyr::filter(., final_data == "Yes") %>% 
  # sort by ID
  dplyr::arrange(., participant_id) %>% 
  # add a samplename column that matches the rest of my data
  dplyr::mutate(., samplename = paste0(
    "STP", stringr::str_pad(participant_id, width = 4, pad = "0"))
    ) %>%
  # ensure fetal sex is a character
  dplyr::mutate(f22_sex = as.character(f22_sex)) %>%
  # make samplename the first column
  dplyr::select(., samplename, everything())

# STOP samples with miRNA plasma
STOP_plasma <- c("STP0016", "STP0023", "STP0105", "STP0148", "STP0230", "STP0275", "STP0282", "STP0378", "STP0582", "STP0583", "STP0593", "STP0596", "STP0636", "STP0677", "STP0692", "STP0761", "STP0788", "STP0795", "STP0798", "STP0852", "STP0867", "STP0888", "STP0903", "STP0912", "STP0932", "STP0944", "STP1002", "STP1082", "STP1087", "STP1206")

# filter the metadata to only inlcude STOP samples with plasma miRNA
subsetSTOP %<>% dplyr::filter(., samplename %in% STOP_plasma)
```

## Convert STOP variables to match SCOPE
- Height is measured in m. Recode to cm to match all the other measurements and SCOPE.    

```{r}
# Convert the height in M to height in cm
subsetSTOP %<>% dplyr::mutate(f9_hgt_cm = f9_hgt * 100)

```

# Combine SCOPE and STOP

```{r combine scope and stop}

combinedMetaData <- data.frame(samplename = c(subsetSCOPE[,"samplename"], subsetSTOP[,"samplename"]),
                               cohort = c(rep("SCOPE", nrow(subsetSCOPE)), rep("STOP", nrow(subsetSTOP))),
                               gestationDelivery = c(subsetSCOPE[,"f39c_final_del_gest"], subsetSTOP[,"f22_gestw_del"]),
                               neonatalSex = c(subsetSCOPE[, "f26_Sex_MF"], subsetSTOP[, "f22_sex"]),
                               neonatalBirthweight = c(subsetSCOPE[, "f26_Birthwgt"], subsetSTOP[, "f22_birthwgt"]),
                               maternalWaist = c(subsetSCOPE[, "f11_waist"], subsetSTOP[, "f9_waist"]),
                               maternalHip = c(subsetSCOPE[, "f11_hip"], subsetSTOP[, "f9_hip"]),
                               maternalBMI = c(subsetSCOPE[, "f11_bmi"], subsetSTOP[, "f9_bmi"]),
                               maternalWgt = c(subsetSCOPE[, "f11_wgt"], subsetSTOP[, "f9_wgt"]),
                               maternalHgt = c(subsetSCOPE[, "f11_hgt"], subsetSTOP[, "f9_hgt_cm"]),
                               maternalAge = c(subsetSCOPE[, "f1_age"], subsetSTOP[, "f1_age"]),
                               maternalBurger_1st = c(subsetSCOPE[, "f10_burger_1st_vst"], subsetSTOP[, "f8_burgers_1st_trim"]),
                               maternalBurger_prepreg = c(subsetSCOPE[, "f10_burger_1m_prepreg"], subsetSTOP[, "f8_burgers_prepreg"]),
                               maternalCurry_1st = c(subsetSCOPE[, "f10_curry_1st_vst"], subsetSTOP[, "f8_curry_1st_trim"]),
                               maternalCurry_prepreg = c(subsetSCOPE[, "f10_curry_1m_prepreg"], subsetSTOP[, "f8_curry_prepreg"]),
                               maternalFriedChicken_1st = c(subsetSCOPE[, "f10_fried_chicken_1st_vst"], subsetSTOP[, "f8_chicken_1st_trim"]),
                               maternalFriedChicken_prepreg = c(subsetSCOPE[, "f10_fried_chicken_1m_prepreg"], subsetSTOP[, "f8_chicken_prepreg"]),
                               maternalFruit_1st = c(subsetSCOPE[, "f10_fruit_1st_vst"], subsetSTOP[, "f8_fruit_1st_trim"]),
                               maternalFruit_prepreg = c(subsetSCOPE[, "f10_fruit_1m_prepreg"], subsetSTOP[, "f8_fruit_prepreg"]),
                               materalLeafyVege_1st = c(subsetSCOPE[, "f10_leafy_vege_1st_vst"], subsetSTOP[, "f8_leafy_vege_1st_trim"]),
                               maternalLeafyVeg_prepreg = c(subsetSCOPE[, "f10_leafy_vege_1m_prepreg"], subsetSTOP[, "f8_leafy_vege_prepreg"]),
                               maternalChips_1st = c(subsetSCOPE[, "f10_chips_1st_vst"], subsetSTOP[, "f8_chips_1st_trim"]),
                               maternalChips_prepreg = c(subsetSCOPE[, "f10_chips_1m_prepreg"], subsetSTOP[, "f8_chips_prepreg"]),
                               maternalOilyFish_1st = c(subsetSCOPE[, "f10_oily_fish_1st_vst"], subsetSTOP[, "f8_oily_fish_1st_trim"]),
                               maternalOilyFish_prepreg = c(subsetSCOPE[, "f10_oily_fish_1m_prepreg"], subsetSTOP[, "f8_oily_fish_prepreg"]),
                               maternalOtherFish_1st = c(subsetSCOPE[, "f10_oth_fish_1st_vst"], subsetSTOP[, "f8_oth_fish_1st_trim"]),
                               maternalOtherFish_prepreg = c(subsetSCOPE[, "f10_oth_fish_1m_prepreg"], subsetSTOP[, "f8_oth_fish_prepreg"]),
                               maternalPizza_1st = c(subsetSCOPE[, "f10_pizza_1st_vst"], subsetSTOP[, "f8_pizza_1st_trim"]),
                               maternalPizza_prepreg = c(subsetSCOPE[, "f10_pizza_1m_prepreg"], subsetSTOP[, "f8_pizza_prepreg"]))


```

# Add outcomes data
- Go back to the metadata and add information regarding pregnancy outcomes.  
```{r}

temp <- dplyr::filter(metadata_quality, samplename %in% c(SCOPE_plasma, STOP_plasma))
```

